# Severe weather effects on health and economy

## Synopsis

## Data processing

### Libraries

First we need to load libraries that we will use later.

```{r, cache=TRUE}
library(ggplot2)
library(plyr)
library(stringr)
library(magrittr)
library(reshape2)
```

### Getting and reading data

The  U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database file is downloaded from the Course web site, but only if it doesn't exist. Then we read the comma-separated-value format. 

```{r, cache=TRUE}

data_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
filename <- "repdata_data_StormData.csv.bz2"

if (!file.exists(filename)){
	download.file(data_url,filename, method="curl")
}

data <- read.csv(bzfile(filename))

```

### Tidying data

The event types are made up of `r length(unique(data$EVTYPE))` different strings. 

```{r}
head(sort(unique(data$EVTYPE)),20)
```

As we can see, there are many spelling mistakes (for example "AVALANCE" instead of "AVALANCHE"), abbreviations, untrimmed strings and unnecessarily detailed classes, so we need to fix that. First we make all values upper case, then change "/" and "&" to "AND" to make it easier on the eyes. We focus on the classes that are the most deadly.

```{r}

# Make column names lower case
colnames(data) <- colnames(data) %>% tolower()

# Make all categories upper case so that "Early Frost" and "EARLY FROST" is the same category.
# Replace & or / with AND. 
data$evtype <- data$evtype %>% toupper() %>% str_trim() %>% sub("[ ]*[$/][ ]*", " AND ", .)

data$evtype[data$evtype == "AVALANCE"] <- "AVALANCHE"

data$evtype <- data$evtype %>% sub("URBAN/SML STREAM FLDG", "URBAN/SMALL STREAM FLOODING",.)
data$evtype <- data$evtype %>% sub("WIND CHILL", "WINDCHILL",.)

# Group Thunderstorms together

data$evtype <- data$evtype %>% sub("TSTM", "THUNDERSTORM", .)
data$evtype <- data$evtype %>% sub("THUNDERSTORMW", "THUNDERSTORM WINDS", .)
data$evtype <- data$evtype %>% sub("THUDERSTORM", "THUNDERSTORM", .)
data$evtype <- data$evtype %>% sub("TUNDERSTORM", "THUNDERSTORM", .)
data$evtype <- data$evtype %>% sub("TUNDERSTORMS", "THUNDERSTORM", .)

data$evtype <- data$evtype %>% sub("TUNDERSTORM W INDS", "THUNDERSTORM WINDS", .)

data$evtype[grepl("THUNDER.*WIND", data$evtype ) & !grepl("HAIL", data$evtype ) & !grepl("LIGHTNING", data$evtype ) & !grepl(
	"FLOOD",data$evtype) & !grepl("TREE",data$evtype) & !grepl("AWNING", data$evtype)] <- "THUNDERSTORM WINDS"

data$evtype <- data$evtype %>% sub("UNSEASONABLE", "UNSEASONABLY",.)

# Remove names for tropical storms

data$evtype[grepl("TROPICAL STORM", data$evtype)] <- "TROPICAL STORM"

data$evtype <- data$evtype %>% sub("TORNADOES", "TORNADO", .)
data$evtype <- data$evtype %>% sub("TORNDAO", "TORNADO", .)

# Tornado and or waterspout
data$evtype[grepl("^TORNADO F[0-5]", data$evtype)] <- "TORNADO"

data$evtype[grepl("WATER.*SPOUT", data$evtype) &  grepl("TORNADO", data$evtype)] <- "TORNADO/WATERSPOUT"
data$evtype[grepl("WATER.*SPOUT", data$evtype) &  !grepl("TORNADO", data$evtype) & !grepl("DUST",data$evtype)] <- "WATERSPOUT"

# Remove "summary of may 21 " types of events
data <- data[!grepl("^SUMMARY", data$evtype),]

data$evtype[grepl("^HAIL.[0-5]*", data$evtype)] <- "HAIL"
data$evtype[grepl("^HIGH WIND.*[0-5]*", data$evtype)] <- "HIGH WIND"

data$evtype[grepl("^WILD.*FIRE", data$evtype)] <- "WILDFIRE"

data$evtype[data$evtype == "RIP CURRENT"] <- "RIP CURRENTS"
data$evtype[data$evtype %in% c("EXTREME HEAT", "EXCESSIVE HEAT", "HEAT WAVE", "RECORD HEAT")] <- "HEAT"
data$evtype[data$evtype == "DENSE FOG"] <- "FOG"
data$evtype[data$evtype == "STRONG WIND"] <- "WIND"
data$evtype[data$evtype == "EXTREME COLD AND WINDCHILL"] <- "COLD AND WINDCHILL"
data$evtype[data$evtype == "HEAVY SURF"] <- "HIGH SURF"
data$evtype[data$evtype == "URBAN AND SML STREAM FLD"] <- "FLOOD"
data$evtype[data$evtype %in% c("WINTER WEATHER AND MIX", "WINTRY MIX", "WINTRY WEATHER MIX")] <- "WINTER WEATHER"

data$evtype[grepl("FLOOD",data$evtype) & !grepl("THUNDER", data$evtype) &  !grepl("RAIN", data$evtype) & !grepl("WIND", data$evtype) & !grepl("LANDSLIDE", data$evtype)] <- "FLOOD"
```

The event types are now made up of `r length(unique(data$evtype))` different strings. Would be nicer to have them strings properly capitalized.

```{r}
# from ?toupper

simpleCap <- function(x) {
    s <- strsplit(x, " ")[[1]]
    paste(toupper(substring(s, 1, 1)), substring(s, 2),
          sep = "", collapse = " ")
}

data$evtype <- data$evtype %>% tolower() %>% sapply(., simpleCap)

```

## Most harmful events with respect to population health

We would like to find the most harmful event types with respect to human health. We sum together injuries and fatalities, so we disregard the seriousness of fatalities versus injuries.

```{r, cache=TRUE, fig.width=12}

a <- aggregate(data=data, cbind(fatalities,injuries)~evtype,FUN=sum)

# Sum fatalities and injuries
a <- transform(a, sum=injuries+fatalities)

# Sort by sum
a <- a[order(-a$sum),]

# Remove sum
a$sum <- NULL

# Make the factor reflect the order
a$evtype <- factor(a$evtype, a$evtype)

# Pick out the 50 worst event types
a <- head(a,50)

# Melt so that we can color after injuries and fatalities
a <- melt(a, id.vars="evtype")

ggplot(data=a, aes(x = evtype, y = value, fill=variable)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + scale_y_log10() 
```

# Results

